---
title: "STA 108 Discussion 1: R Basics"
author: "Yuanyuan Li (yynli@ucdavis.edu)"
output:
  pdf_document: default
  html_document:
    df_print: paged
urlcolor: blue
---

```{r, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(
	error = FALSE,
	message = FALSE,
	warning = FALSE
)
knitr::opts_chunk$set(fig.width=5, fig.height=3) 
```

# Installation
##1. Install R:
1. Go to http://cran.freestatistics.org/.
2. Select the appropriate package based on your operating system.
3. Follow the installers instruction.

##2. Install RStudio:
1. Go to https://rstudio.com/products/rstudio/download/.
2. Select the “Desktop” version.
3. Select the installer based on your operating system (usually the recommended for your system works
just fine).

##3. Install the R Markdown packages (Recommended for class projects):
1. Install Markdown package by typing this command in R **Console** (I will explain it in next section):

```{r,eval=FALSE}
install.packages("rmarkdown")
```

2. Install LaTeX (TinyTeX) for PDF reports by typing this command in R **Console**:
```{r,eval=FALSE}
install.packages("tinytex")
```

# Starting out

* Click RStudio icon to run this program, and R will be open automatically in the background at the same time. You should have 4 panels. A brief description of the four windows follows:
    - **Environment, History, Build, VCS**: This window lists all Data, Values Functions entered.
    - **Files, Plots, Packages, Help**: Should you use the help command (the ?) in R, the relevant help document will be opened here.Should you plot a figure, the relevant figure will be displayed.
    - **Console**: Where you type in your commands and retrieve your output.
    - **Source**: A  convenient  place  to  store  your  commands,  which  can  then  be  sent  to  the  console  via keyboard shortcuts or with a button. I **highly** recommend putting all of your code in the source window (in a R Script or R markdown), and transferring it to the console.


* Ways of typing commands in R:
    + Type code in Console
    + Create new script file and type code in editor (**recommended**)

* Running R code:
    - In R Console: hit *Enter* after the command line
    - In R Script editor: Highlight the part you need to run and hit     the "run" symbol button on the menu; or apply the shortcut: *Ctrl/command + Return*(MAC) or *Ctrl + Enter*(Windows)

* Change working directory:
    - go to Session > Set Working Directory > to source file location or Choose Directory
    - Or use "setwd("file-path")" command

* Save R Source files:
    - click the blue square button 
    - click File > Save/ Save as

Let's start coding!

# 1. Calculation: Use R as a calculator

```{r}
31/4*(37-25) 
3^2
sqrt(36)
log(4)
cos(6)
```

# 2. Objects 
## Assign values to object 'x' using any one of the following:
```{r}
x = 5 
x <- 5
5 -> x
```


## Calculation
```{r}
x+3
x^2
sqrt(x)
y = x^2
```

## Overwrite the existing objects
```{r}
x = 10
x = x+1
```


#3. Vectors
```{r}
# Create a vector
c(1,3,2,4)
# Save the vector as 'x'
x = c(1,3,2,4)

# R applies functions to every element of a vector
x - 10
x^2
```


#4. Some useful functions
```{r}
mean(x) # mean
sd(x) # standard deviation
var(x) # variance
summary(x)
sum(x) # sum of all elements
prod(x) # product of all elements
length(x) # number of elements

x[1:3] # the first three elements
```

#5. Import the data
* Read in a data set by specifying the full file path (remember to replace “\\” with “/” if your file path include “\\”): 
```{r}
patients101 = read.csv("~/books/108s21/Discussion/datasets/patients101.csv")
```
    

* Read in a data set by setting the parent folder as working directory, then use "read.csv" or "read.table" function:
```{r}
setwd("~/books/108s21/Discussion/datasets")#set working directory to "datasets" folder
patients101 = read.csv("patients101.csv")#read the data set in the folder
```

* Use "Import Dataset" button in the menu of Environment: 
    + Read .txt file or .csv file : Import Dataset>From Text(base)>choose file>open>import 
    + Read Excel file : Import Dataset>From Excel>choose file>open>import 
    
#6. Accessing specific rows and columns in a dataframe
```{r}
head(patients101)#Display the first six rows
patients101[1:6,]#Display the first six rows by row index
Y = patients101$sysBP # Extract variables from dataset
Y = patients101[,3] # or Extract variables by column index
patients101[1,3]#Display the value is row 1 and column 3
```


#7. Plot 
```{r}
# histograms
hist(Y) 
hist(Y, xlab = 'sysBP', ylab = 'frequency', main = 'Histogram of sysBP')

# boxplots
boxplot(Y, main = 'Boxplot of sysBP')

# scatterplots
X1 = patients101$weight
plot(X1, Y, xlab = 'weight', ylab = 'sysBP', main = 'Plot of weight versus sysBP')
```



