---
title: "R markdown basics"
author: "Yuanyuan Li"
output: pdf_document
urlcolor: blue
---


#Introduction
## Why use R markdown?
1. Markdown is a simple formatting syntax for authoring **HTML, PDF, and MS Word** documents. 

2. Markdown can combine R codes, results, graph with words in a neat way.

3. Markdown can create beautiful mathmatical symbols and equations using LaTeX Language (r package: *TinyTeX*), create tables (r package: *knitr*).


##Create a R markdown file
1. Make sure you already have *rmarkdown* package (check discussion1.pdf for installing it).

2. In RStudio, click the green "+" in the top left> R markdown > write a title and choose output file format > OK

3. Type text in the white space for your report, and only write R code in the grey R chunck, which can be created by click Insert>R or type ` ```{r} ` at the begining and ` ``` ` at the end.

4. When you click the **Knit** button, a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.

#Links
This is a link for introducing R Markdown Basics:
https://rmarkdown.rstudio.com/authoring_basics.html

A R Markdown Cookbook: https://bookdown.org/yihui/rmarkdown-cookbook/r-code.html


#R Code Chunks
You can embed an R code chunk like this:

```{r}
summary(cars)
```


**Recommend**: If you want to hide the R code chunk, and only show the results, you can use  `echo = FALSE` parameter in the R code chunk:

```{r, echo=FALSE}
summary(cars)
```

#Inline R code
There were `r nrow(cars)` cars studied.

#Size of plots/images

##1. Set the size of plots globally

Copy the R chunk below and paste at the begining of your RMD file, use ` ```{r, include=FALSE}``` ` to hide this chunk:

```{r}
knitr::opts_chunk$set(fig.width=4, fig.height=4)
```


```{r}
hist(cars$dist,main="Histogram of distance",xlab="distance")
```

##2. Set the size of plot for one chunk
We can set the chunk options for each chunk too. Use ` {r, fig.height = 3, fig.width = 5} ` as in below chunk:

```{r, fig.height = 3, fig.width = 6}
hist(cars$dist,main="Histogram of distance",xlab="distance")
```

#Create tables

Use `knitr::kable` function on an R object, which is typically a matrix or data frame. Type "?kable" in R console to get its help document.

```{r}
mydata=head(mtcars)
knitr::kable(mydata)
```


\section{Section titles}
\subsection{Section titles}
\subsubsection{Section titles}
#Section titles
##Section titles
###Section titles

#Lists
##Unordered lists
- Item 1
- Item 2
    - Item 2a
    - Item 2b

##Ordered List:
1. Item 1
2. Item 2
3. Item 3
    - Item 3a
    - Item 3b
    


#LaTeX Equations
$2+4$
$\alpha$

The formula for mean is $\bar{x}=\frac{1}{n}\sum_{i=1}^n x_i$.
Display the formula in a new line:
$$\bar{x}=\frac{1}{n}\sum_{i=1}^n x_i$$

#Appendix
```{r, ref.label=knitr::all_labels(),echo=TRUE,eval=FALSE}
```
